Visualisation of Brain Statistics with R-packages \emph{ggseg} and \emph{ggseg3d}

### 2020-03-17 msc_ggseg.tex
update with revisions comments

In this tutorial, we introduce two packages for visualizing brain atlas data in R.
The \emph{ggseg} and \emph{ggseg3d} -- pluss the complimentary \emph{ggsegExtra} -- packages include pre-compiled data sets for different brain atlases that allow for 2D and 3D visualization.
The two-dimensional functionality in \emph{ggseg} is based on polygons and \emph{ggplot2}-based grammar of graphics (\citet{ggplot}), while the 3D functionality in \emph{ggseg3d} is based on tri-surface mesh plots and \emph{plotly} (\citet{plotly}).

Both packages present compiled data sets, tailored functions that allow brain data integration and plotting, and other minor features such as custom colour palettes.
The data featured in the packages are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DKT; \citet{dkt}), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; \citet{aseg}), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (\citet{fischl_99}, \citet{dale_99}, \citet{Fischl2000}), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (\citet{amlien_elaboration_2019}, \citet{WALHOVD20051261}, \citet{Pizzagalli}).
The \emph{ggsegExtra} package contains a collection of precompiled atlases (currently 15 additional atlases) and it is frequently updated.
A summary of all available atlases compatible with the \emph{ggseg}-packages as of December 2019 can be found in Table \ref{tab:atlasTab}.

@@

\caption{\label{tab:atlasTab}Table of currently available atlases in the ggseg, ggseg3d, and the ggsegExtra R-packages. Polygon and mesh refer to 2D and 3D brain atlas representations, respectively}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllll}
\toprule
Title & Item & Mesh & Polygon & citation\\
\midrule
Automated probabilistic reconstruction of white-matter pathways & tracula & None & ggsegExtra & \citet{tracula}\\
Desikan-Killiany Cortical Atlas & dkt & ggseg3d & ggseg & \citet{dkt}\\
Desterieux cortical parcellations & desterieux & ggsegExtra & None & \citet{desterieux}\\
Freesurfer automatic subcortical segmentation of a brain volume & aseg & ggseg3d & ggseg & \citet{aseg}\\
Genetic topography of brain area morphology & chenAr & None & ggsegExtra & \citet{chen}\\
\addlinespace
Genetic topography of brain thickness morphology & chenTh & None & ggsegExtra & \citet{chen}\\
Harvard-Oxford Cortical atlas & hoCort & None & ggsegExtra & \citet{ho}\\
HPC - Multi-modal parcellation of human cerebral cortex & glasser & None & ggsegExtra & \citet{glasser}\\
ICBM white matter parcellation & icbm & ggsegExtra & None & \citet{icbm}\\
JHU parcellation & jhu & None & ggsegExtra & \citet{jhu}\\
\addlinespace
Local-Global 17 Parcellation of the Human Cerebral Cortex & schaefer17 & ggsegExtra & None & \citet{schaefer}\\
Local-Global 7 Parcellation of the Human Cerebral Cortex & schaefer7 & ggsegExtra & None & \citet{schaefer}\\
Parcellation from JHU & jhu & ggsegExtra & None & \citet{jhu}\\
Parcellation from the Human Connectome Project & glasser & ggsegExtra & None & \citet{glasser}\\
White matter tract parcellations & tracula & ggsegExtra & None & \citet{tracula}\\
\addlinespace
Yeo 17 Resting-state Cortical Parcellations & yeo17 & ggsegExtra & ggsegExtra & \citet{yeo2011}\\
Yeo 7 Resting-state Cortical Parcellations & yeo7 & ggsegExtra & ggsegExtra & \citet{yeo2011}\\
\bottomrule
\end{tabular}}
\end{table}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# remotes::install\_github("LCBC{-}UiO/ggseg")}
\KeywordTok{library}\NormalTok{(ggseg)}
\KeywordTok{library}\NormalTok{(dplyr)}
\KeywordTok{library}\NormalTok{(tidyr)}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# dkt dark theme}
\NormalTok{p1 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{position =} \StringTok{"stacked"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme\_dark}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{" "}\NormalTok{)}

\CommentTok{\# dkt classic theme}
\NormalTok{p2 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{position =} \StringTok{"stacked"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme\_classic}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{\# dkt medial view}
\NormalTok{med <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{view =} \StringTok{"medial"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{\# dkt left hemisphere}
\NormalTok{left <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{hemisphere =} \StringTok{"left"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{\# aseg default theme}
\NormalTok{p3 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{atlas=}\NormalTok{aseg) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{\# dkt left medial alone}
\NormalTok{combo <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{view =} \StringTok{"medial"}\NormalTok{,}
               \DataTypeTok{hemisphere =} \StringTok{"left"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{" "}\NormalTok{)}

\CommentTok{\# Combine plots to Figure 2}
\NormalTok{cowplot}\OperatorTok{::}\KeywordTok{plot\_grid}\NormalTok{(p1, med, combo,  p2, left, p3,}
                   \DataTypeTok{labels =} \KeywordTok{c}\NormalTok{(}\StringTok{"A: dkt {-} dark"}\NormalTok{, }\StringTok{"B: dkt {-} medial"}\NormalTok{,}
                              \StringTok{"C: dkt {-} combo"}\NormalTok{, }\StringTok{"D: dkt {-} classic"}\NormalTok{,}
                              \StringTok{"E: dkt {-} left"}\NormalTok{, }\StringTok{"F: aseg"}\NormalTok{),}
                   \DataTypeTok{hjust =} \FloatTok{{-}.05}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{msc_ggseg_files/figure-latex/collection-1.pdf}
\caption{\label{fig:collection}\emph{ggseg} plots can be used with any \emph{ggplot2} feature such as standard scales and themes. For cortical atlases, one can supply special \emph{ggseg} options to determine, for example, hemisphere, view, or position. \textbf{A:} \texttt{dkt} atlas, stacked with dark theme ; \textbf{B:} \texttt{dkt} with medial view only; \textbf{C:} \texttt{dkt} atlas with only left medial display; \textbf{D:} \texttt{dkt} atlas, stacked, with classic theme; \textbf{E:} \texttt{dkt} atlas with left hemisphere only; \textbf{F:} complete \texttt{aseg} atlas}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Figure 3}
\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{mapping=}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill =}\NormalTok{ area), }\DataTypeTok{colour=}\StringTok{"black"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale\_fill\_brain}\NormalTok{(}\StringTok{"dkt"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.justification=}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{0}\NormalTok{),}
        \DataTypeTok{legend.position=}\StringTok{"bottom"}\NormalTok{,}
        \DataTypeTok{legend.text =} \KeywordTok{element\_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{5}\NormalTok{)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{guides}\NormalTok{(}\DataTypeTok{fill =} \KeywordTok{guide\_legend}\NormalTok{(}\DataTypeTok{ncol =} \DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

In any atlas, the column `label' is particularly useful for combining the data of interest with the \emph{ggseg}-polygons.
The column `label' contains the label (region) names as in the original neuroimaging software.
For example, the DKT atlas label column matches the region names from Freesurfer statistics table outputs.
Yet, the data in \emph{ggseg} is in a long format - that is each region has a single row - and any data of interest needs to be in this same format.
Often data sets are organized in wide format, in which subjects are represented by rows and each different data variable is represented in a separate column, and thus need to be rearranged to work with ggseg.
See below an example of wide-to-long conversion.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Create a mock freesurfer output}
\NormalTok{freesurfer\_stats <{-}}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}
  \DataTypeTok{id =} \KeywordTok{c}\NormalTok{(}\DecValTok{10}\OperatorTok{:}\DecValTok{12}\NormalTok{),}
  \DataTypeTok{lh\_superiortemporal =} \KeywordTok{c}\NormalTok{(}\FloatTok{3.32}\NormalTok{, }\FloatTok{4.1}\NormalTok{, }\FloatTok{3.5}\NormalTok{),}
  \DataTypeTok{lh\_precentral =} \KeywordTok{c}\NormalTok{(}\FloatTok{2.3}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\FloatTok{2.1}\NormalTok{),}
  \DataTypeTok{lh\_rostralmiddlefrontal =} \KeywordTok{c}\NormalTok{(}\FloatTok{3.3}\NormalTok{, }\FloatTok{3.2}\NormalTok{, }\FloatTok{3.1}\NormalTok{)}
\NormalTok{)}
\NormalTok{freesurfer\_stats}

\CommentTok{\# Wrangle wide format data into long format}
\NormalTok{freesurfer\_long <{-}}\StringTok{ }\NormalTok{freesurfer\_stats }\OperatorTok{\%>\%}
\StringTok{  }\KeywordTok{gather}\NormalTok{(label, thickness, }\OperatorTok{{-}}\NormalTok{id)}
\NormalTok{freesurfer\_long}
\CommentTok{\#\#   id lh\_superiortemporal lh\_precentral lh\_rostralmiddlefrontal}
\CommentTok{\#\# 1 10                3.32           2.3                     3.3}
\CommentTok{\#\# 2 11                4.10           2.5                     3.2}
\CommentTok{\#\# 3 12                3.50           2.1                     3.1}
\CommentTok{\#\#   id                   label thickness}
\CommentTok{\#\# 1 10     lh\_superiortemporal      3.32}
\CommentTok{\#\# 2 11     lh\_superiortemporal      4.10}
\CommentTok{\#\# 3 12     lh\_superiortemporal      3.50}
\CommentTok{\#\# 4 10           lh\_precentral      2.30}
\CommentTok{\#\# 5 11           lh\_precentral      2.50}
\CommentTok{\#\# 6 12           lh\_precentral      2.10}
\CommentTok{\#\# 7 10 lh\_rostralmiddlefrontal      3.30}
\CommentTok{\#\# 8 11 lh\_rostralmiddlefrontal      3.20}
\CommentTok{\#\# 9 12 lh\_rostralmiddlefrontal      3.10}
\end{Highlighting}
\end{Shaded}

@@

Data in long format can then be used directly with the \texttt{ggseg()}-function, as the `label' column corresponds in name and content with the `label' column in the atlas data of \texttt{dkt}.
The data \textbf{must} include a column that has the same column name and at least \emph{some} data matching the values in the corresponding column in the atlas data.
In the next example we create some data with 4 rows, with an `area' and `p' column, representing the results of a hypothetical analysis.
The \texttt{ggseg()}-function will recognise the matching column `area', and merge the supplied data into the atlas using \emph{dplyr}-joins.
We use the `p' column as the column flooding the segment with colour.
The appearance of the plot can then be modified similarly to any other \emph{ggplot2} graph using functions such as scales, labs, themes, etc., as seen in Figure \ref{fig:datasupp}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Make some mock data}
\NormalTok{someData =}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}
  \DataTypeTok{area =} \KeywordTok{c}\NormalTok{(}\StringTok{"transverse temporal"}\NormalTok{, }\StringTok{"insula"}\NormalTok{,}
           \StringTok{"pre central"}\NormalTok{,}\StringTok{"superior parietal"}\NormalTok{),}
  \DataTypeTok{p =} \KeywordTok{sample}\NormalTok{(}\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{,.}\DecValTok{001}\NormalTok{), }\DecValTok{4}\NormalTok{),}
  \DataTypeTok{stringsAsFactors =} \OtherTok{FALSE}\NormalTok{)}

\CommentTok{\# Figure 4}
\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{.data=}\NormalTok{someData, }\DataTypeTok{mapping=}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill=}\NormalTok{p))  }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"A nice plot title"}\NormalTok{, }\DataTypeTok{fill=}\StringTok{"p{-}value"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale\_fill\_gradient}\NormalTok{(}\DataTypeTok{low=}\StringTok{"firebrick"}\NormalTok{,}\DataTypeTok{high=}\StringTok{"goldenrod"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

>>>>

In this tutorial, we introduce two packages for visualizing brain atlas data in R.
The \emph{ggseg} and \emph{ggseg3d} -- plus the complimentary \emph{ggsegExtra} -- packages include pre-compiled data sets for different brain atlases that allow for 2D and 3D visualization.
The two-dimensional functionality in \emph{ggseg} is based on polygons and \emph{ggplot2}-based grammar of graphics (\citet{ggplot}), while the 3D functionality in \emph{ggseg3d} is based on tri-surface mesh plots and \emph{plotly} (\citet{plotly}).

Both packages present compiled data sets, tailored functions that allow brain data integration and plotting, and other minor features such as custom colour palettes.
The data featured in the packages are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DKT; \citet{dkt}), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; \citet{aseg}), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (\citet{fischl_99}, \citet{dale_99}, \citet{Fischl2000}), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (\citet{amlien_elaboration_2019}, \citet{WALHOVD20051261}, \citet{Pizzagalli}).
The \emph{ggsegExtra} package contains a collection of precompiled atlases (currently 16 additional atlases) and it is frequently updated.
A summary of all available atlases compatible with the \emph{ggseg}-packages as of March 2020 can be found in Table \ref{tab:atlasTab}.

@@

\caption{\label{tab:atlasTab}Table of currently available atlases in the ggseg, ggseg3d, and the ggsegExtra R-packages. Polygon and mesh refer to 2D and 3D brain atlas representations, respectively}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllll}
\toprule
Title & Item & Mesh & Polygon & Citation\\
\midrule
Automated probabilistic reconstruction of white-matter pathways & tracula & None & ggsegExtra & \citet{tracula}\\
Desikan-Killiany Cortical Atlas & dkt & ggseg3d & ggseg & \citet{dkt}\\
Desterieux cortical parcellations & desterieux & ggsegExtra & None & \citet{desterieux}\\
Freesurfer ASEG with posterior-anterior hippocampus & hcpa & ggsegExtra & None & \\
Freesurfer automatic subcortical segmentation of a brain volume & aseg & ggseg3d & ggseg & \citet{aseg}\\
\addlinespace
Genetic topography of brain area morphology & chenAr & None & ggsegExtra & \citet{chen}\\
Genetic topography of brain thickness morphology & chenTh & None & ggsegExtra & \citet{chen}\\
Harvard-Oxford Cortical atlas & hoCort & None & ggsegExtra & \citet{ho}\\
HPC - Multi-modal parcellation of human cerebral cortex & glasser & None & ggsegExtra & \citet{glasser}\\
ICBM white matter parcellation & icbm & ggsegExtra & None & \citet{icbm}\\
\addlinespace
JHU parcellation & jhu & None & ggsegExtra & \citet{jhu}\\
Local-Global 17 Parcellation of the Human Cerebral Cortex & schaefer17 & ggsegExtra & None & \citet{schaefer}\\
Local-Global 7 Parcellation of the Human Cerebral Cortex & schaefer7 & ggsegExtra & None & \citet{schaefer}\\
Parcellation from JHU & jhu & ggsegExtra & None & \citet{jhu}\\
Parcellation from the Human Connectome Project & glasser & ggsegExtra & None & \citet{glasser}\\
\addlinespace
White matter tract parcellations & tracula & ggsegExtra & None & \citet{tracula}\\
Yeo 17 Resting-state Cortical Parcellations & yeo17 & ggsegExtra & ggsegExtra & \citet{yeo2011}\\
Yeo 7 Resting-state Cortical Parcellations & yeo7 & ggsegExtra & ggsegExtra & \citet{yeo2011}\\
\bottomrule
\end{tabular}}
\end{table}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# remotes::install\_github("LCBC{-}UiO/ggseg")}
\KeywordTok{library}\NormalTok{(ggseg)}

\CommentTok{\# package combining plots easily}
\KeywordTok{library}\NormalTok{(patchwork)}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# A: dkt dark theme}
\NormalTok{p1 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{position =} \StringTok{"stacked"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme\_dark}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"dkt"}\NormalTok{, }\DataTypeTok{subtitle =} \StringTok{"dark theme"}\NormalTok{)}

\CommentTok{\# B: dkt medial view}
\NormalTok{p2 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{view =} \StringTok{"medial"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{"dkt"}\NormalTok{, }\DataTypeTok{subtitle =} \StringTok{"medial view"}\NormalTok{)}

\CommentTok{\# C: dkt left medial alone}
\NormalTok{p3 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{view =} \StringTok{"medial"}\NormalTok{,}
            \DataTypeTok{hemisphere =} \StringTok{"left"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"dkt"}\NormalTok{, }\DataTypeTok{subtitle =} \StringTok{"medial left side"}\NormalTok{)}

\CommentTok{\# D: dkt classic theme}
\NormalTok{p4 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{position =} \StringTok{"stacked"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme\_classic}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{"dkt"}\NormalTok{, }\DataTypeTok{subtitle =} \StringTok{"classic theme"}\NormalTok{)}

\CommentTok{\# E: dkt left hemisphere}
\NormalTok{p5 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{hemisphere =} \StringTok{"left"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{"dkt"}\NormalTok{, }\DataTypeTok{subtitle =} \StringTok{"left hemisphere"}\NormalTok{)}

\CommentTok{\# F: aseg default theme}
\NormalTok{p6 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{atlas =}\NormalTok{ aseg, }\DataTypeTok{view =} \StringTok{"axial"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{"aseg"}\NormalTok{, }\DataTypeTok{subtitle =} \StringTok{"axial view"}\NormalTok{)}

\CommentTok{\# Using patchwork to combine all plots into one}
\NormalTok{(p1 }\OperatorTok{|}\StringTok{ }\NormalTok{p2 }\OperatorTok{|}\StringTok{ }\NormalTok{p3) }\OperatorTok{/}\StringTok{ }\NormalTok{(p4 }\OperatorTok{|}\StringTok{ }\NormalTok{p5 }\OperatorTok{|}\StringTok{ }\NormalTok{p6) }\OperatorTok{+}
\StringTok{  }\KeywordTok{plot\_annotation}\NormalTok{(}\DataTypeTok{tag\_levels =} \StringTok{\textquotesingle{}A\textquotesingle{}}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{msc_ggseg_files/figure-latex/collection-1.pdf}
\caption{\label{fig:collection}\emph{ggseg} plots can be used with any \emph{ggplot2} feature such as standard scales and themes. For cortical atlases, one can supply special \emph{ggseg} options to determine, for example, hemisphere, view, or position. \textbf{A:} \texttt{dkt} atlas, stacked with dark theme ; \textbf{B:} \texttt{dkt} with medial view only; \textbf{C:} \texttt{dkt} atlas with only left medial display; \textbf{D:} \texttt{dkt} atlas, stacked, with classic theme; \textbf{E:} \texttt{dkt} atlas with left hemisphere only; \textbf{F:} axial view of \texttt{aseg} atlas}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Figure 3}
\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill =}\NormalTok{ area), }\DataTypeTok{colour=}\StringTok{"black"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale\_fill\_brain}\NormalTok{(}\StringTok{"dkt"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.justification =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{0}\NormalTok{),}
        \DataTypeTok{legend.position =} \StringTok{"bottom"}\NormalTok{,}
        \DataTypeTok{legend.text =} \KeywordTok{element\_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{5}\NormalTok{)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{guides}\NormalTok{(}\DataTypeTok{fill =} \KeywordTok{guide\_legend}\NormalTok{(}\DataTypeTok{ncol =} \DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

In any atlas, the column `label' is particularly useful for combining the data of interest with the \emph{ggseg}-polygons.
The column `label' contains the label (region) names as in the original neuroimaging software.
For example, the DKT atlas label column matches the region names from Freesurfer statistics table outputs.
Yet, the data in \emph{ggseg} is in a long format - that is each region has a single row - and any data of interest needs to be in this same format.
Often data sets are organized in wide format, in which subjects are represented by rows and each different data variable is represented in a separate column, and thus need to be rearranged to work with ggseg.
To ease to conversion between Freesurfer files and ggseg compatible files, we have functions to aid import to R.
There is an extended vignette in the \href{}{online documentation} of the various options, here we assume that that the data has been compiled into a file using Freesurfer's \texttt{aparcstats2table} or \texttt{asegstats2table} that combine parcellation metrics for all wanted subjects into a single file.
This file also contains extra metrics like intracranial volume and total grey matter.
These do not map onto ggseg, and as such, one should make sure to remove them from the subsequent data.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{file\_path <{-}}\StringTok{ "data/aparc.volume.table"}

\NormalTok{someData <{-}}\StringTok{ }\KeywordTok{read\_freesurfer\_table}\NormalTok{(file\_path, }\DataTypeTok{measure =} \StringTok{"volume"}\NormalTok{) }\OperatorTok{\%>\%}\StringTok{ }
\StringTok{  }\CommentTok{\# remove rows in the data that do not conain the strings "rh" or "lh"}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(}\KeywordTok{grepl}\NormalTok{(}\StringTok{"rh|lh"}\NormalTok{, label))}
\NormalTok{someData}
\CommentTok{\#\# \# A tibble: 34 x 3}
\CommentTok{\#\#    subject label                      volume}
\CommentTok{\#\#    <fct>   <chr>                       <dbl>}
\CommentTok{\#\#  1 bert    rh\_bankssts                  1969}
\CommentTok{\#\#  2 bert    rh\_caudalanteriorcingulate   2280}
\CommentTok{\#\#  3 bert    rh\_caudalmiddlefrontal       5390}
\CommentTok{\#\#  4 bert    rh\_cuneus                    2998}
\CommentTok{\#\#  5 bert    rh\_entorhinal                1735}
\CommentTok{\#\#  6 bert    rh\_fusiform                  8144}
\CommentTok{\#\#  7 bert    rh\_inferiorparietal         14876}
\CommentTok{\#\#  8 bert    rh\_inferiortemporal         11016}
\CommentTok{\#\#  9 bert    rh\_isthmuscingulate          1983}
\CommentTok{\#\# 10 bert    rh\_lateraloccipital         12729}
\CommentTok{\#\# \# ... with 24 more rows}
\end{Highlighting}
\end{Shaded}

@@

A second option is to read in atlas files directly from the entire subject directory, using regular expression to define the atlas wanted.
If we inspect a single subjects stats directory of Freesurfer (here on default installation on macOS), we can see several parcellations to choose from.
The default aparc has a unique string ending with \texttt{aparc.stats} which we will use to read in all subject stats for that particular atlas.
This data has more columns, as it contains several metrics that can be plotted.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# in bash terminal}
\FunctionTok{ls}\NormalTok{ /Applications/freesurfer/subjects/bert/stats}
\CommentTok{\#\# aseg.stats}
\CommentTok{\#\# lh.BA\_exvivo.stats}
\CommentTok{\#\# lh.BA\_exvivo.thresh.stats}
\CommentTok{\#\# lh.aparc.DKTatlas.stats}
\CommentTok{\#\# lh.aparc.a2009s.stats}
\CommentTok{\#\# lh.aparc.pial.stats}
\CommentTok{\#\# lh.aparc.stats}
\CommentTok{\#\# lh.curv.stats}
\CommentTok{\#\# lh.w{-}g.pct.stats}
\CommentTok{\#\# rh.BA\_exvivo.stats}
\CommentTok{\#\# rh.BA\_exvivo.thresh.stats}
\CommentTok{\#\# rh.aparc.DKTatlas.stats}
\CommentTok{\#\# rh.aparc.a2009s.stats}
\CommentTok{\#\# rh.aparc.pial.stats}
\CommentTok{\#\# rh.aparc.stats}
\CommentTok{\#\# rh.curv.stats}
\CommentTok{\#\# rh.w{-}g.pct.stats}
\CommentTok{\#\# wmparc.stats}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{atlas\_data <{-}}\StringTok{ }\KeywordTok{read\_atlas\_files}\NormalTok{(}
  \DataTypeTok{subjects\_dir =} \StringTok{"/Applications/freesurfer/subjects/"}\NormalTok{,}
  \DataTypeTok{atlas =} \StringTok{"aparc.stats$"}
\NormalTok{)}
\NormalTok{atlas\_data}
\CommentTok{\#\# \# A tibble: 204 x 11}
\CommentTok{\#\#    subject label NumVert SurfArea GrayVol ThickAvg}
\CommentTok{\#\#    <chr>   <chr>   <int>    <int>   <int>    <dbl>}
\CommentTok{\#\#  1 bert    lh\_b\textasciitilde{}    1181      831    2297     2.77}
\CommentTok{\#\#  2 bert    lh\_c\textasciitilde{}     843      572    1534     2.72}
\CommentTok{\#\#  3 bert    lh\_c\textasciitilde{}    2758     1840    5772     2.80}
\CommentTok{\#\#  4 bert    lh\_c\textasciitilde{}    2683     1654    3074     1.81}
\CommentTok{\#\#  5 bert    lh\_e\textasciitilde{}     581      416    1840     3.33}
\CommentTok{\#\#  6 bert    lh\_f\textasciitilde{}    4113     2875    8519     2.72}
\CommentTok{\#\#  7 bert    lh\_i\textasciitilde{}    4948     3466   10559     2.70}
\CommentTok{\#\#  8 bert    lh\_i\textasciitilde{}    5056     3542   12358     2.98}
\CommentTok{\#\#  9 bert    lh\_i\textasciitilde{}    1561      990    2350     2.09}
\CommentTok{\#\# 10 bert    lh\_l\textasciitilde{}    7961     5077   12743     2.30}
\CommentTok{\#\# \# ... with 194 more rows, and 5 more variables:}
\CommentTok{\#\# \#   ThickStd <dbl>, MeanCurv <dbl>,}
\CommentTok{\#\# \#   GausCurv <dbl>, FoldInd <int>, CurvInd <dbl>}
\end{Highlighting}
\end{Shaded}

This long format data can be used directly with the \texttt{ggseg()}-function, as the `label' column corresponds in name and content with the `label' column in the atlas data of \texttt{dkt}.
The data \textbf{must} include a column that has the same column name and at least \emph{some} data matching the values in the corresponding column in the atlas data.
In the following example, the \texttt{ggseg()}-function will recognise the matching column `label', and merge the supplied data into the atlas using \emph{dplyr}-joins.
The appearance of the plot can then be modified similarly to any other \emph{ggplot2} graph using functions such as scales, labs, themes, etc., as seen in Figure \ref{fig:datasupp}

\begin{Shaded}
\begin{Highlighting}[]

\CommentTok{\# Figure 4A}
\NormalTok{p1 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{.data=}\NormalTok{someData, }\DataTypeTok{mapping=}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill=}\NormalTok{volume))  }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"from read\_freesurfer\_table"}\NormalTok{, }
       \DataTypeTok{fill=}\StringTok{"volume}\CharTok{\textbackslash{}n}\StringTok{(mm\^{}3)"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale\_fill\_gradient}\NormalTok{(}\DataTypeTok{low=}\StringTok{"firebrick"}\NormalTok{,}\DataTypeTok{high=}\StringTok{"goldenrod"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{guides}\NormalTok{(}\DataTypeTok{fill =} \KeywordTok{guide\_colourbar}\NormalTok{(}\DataTypeTok{barheight =} \DecValTok{3}\NormalTok{))}

\CommentTok{\# Figure 4B}
\NormalTok{p2 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{.data=}\NormalTok{atlas\_data, }\DataTypeTok{mapping=}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill=}\NormalTok{ThickAvg))  }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"from read\_atlas\_files"}\NormalTok{, }
       \DataTypeTok{fill=}\StringTok{"Thickness}\CharTok{\textbackslash{}n}\StringTok{(mean)"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale\_fill\_gradient}\NormalTok{(}\DataTypeTok{low=}\StringTok{"forestgreen"}\NormalTok{,}\DataTypeTok{high=}\StringTok{"goldenrod"}\NormalTok{)}\OperatorTok{+}
\StringTok{  }\KeywordTok{guides}\NormalTok{(}\DataTypeTok{fill =} \KeywordTok{guide\_colourbar}\NormalTok{(}\DataTypeTok{barheight =} \DecValTok{3}\NormalTok{))}

\NormalTok{(p1 }\OperatorTok{/}\StringTok{ }\NormalTok{p2) }\OperatorTok{+}
\StringTok{  }\KeywordTok{plot\_annotation}\NormalTok{(}\DataTypeTok{tag\_levels =} \StringTok{"A"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}
